\pkg{cv}: An \proglang{R} Package for Cross-Validating Regression Models

### 2024-05-23 scratch/notes_john/JSS-article/JSS-article-3.Rmd
add section on other R CV packages to JSS paper

@@

Let $\mathrm{CV}(\mathbf{y}, \widehat{\mathbf{y}})$ represent a cross-validation cost criterion, such as mean-squared error, computed for all of the $n$ values of the response $\mathbf{y}$ based on fitted values $\widehat{\mathbf{y}}$ from the model fit to all of the data. We require that $\mathrm{CV}(\mathbf{y}, \widehat{\mathbf{y}})$ is the mean of casewise components, that is, $\mathrm{CV}(\mathbf{y}, \widehat{\mathbf{y}}) = \frac{1}{n}\sum_{i=1}^n\mathrm{cv}(y_i, \widehat{y}_i)$.[^contrast-function] For example, $\mathrm{MSE}(\mathbf{y}, \widehat{\mathbf{y}}) = \frac{1}{n}\sum_{i=1}^n (y_i - \widehat{y}_i)^2$.

>>>>

The \pkg{cv} package is far from unique in implementing cross-validation of regression models in \proglang{R}. We've already mentioned the \code{cv.glm()} function in the \pkg{boot} package. A general review of CV facilities in \proglang{R} is beyond the scope of the current paper, but in this section we remark selectively on other \proglang{R} packages that support CV.

Our goal in implementing CV is to make it conveniently available to \proglang{R} users with limited programming skills,[^cv-by-programming] and hence to encourage its use. Towards this end, the \pkg{cv} package provides a simple, consistent interface through the \code{cv()} generic function, with specific methods for different classes of standard \proglang{R} statistical models. Moreover, we have tried to make writing methods for additional classes of statistical models as simple and straightforward as possible. Because CV is computationally intensive, we also aim for efficiency, by enabling parallel computations generally, and by exploiting properties of specific classes of statistical models (in particular, computations based on hatvalues and the Woodbury matrix identity for linear and generalized linear model). Finally, the package includes some unusual features, such as general support for mixed-effects models and for cross-validating complex model-specification procedures.

[^cv-by-programming]: \proglang{R} users with sophisticated programming skills would generally find it unchallenging to implement CV directly for specific applications. Even in this case, however, there's an argument for using a simple pre-programmed interface to CV, to minimize programming effort and to avoid mistakes.

Some \proglang{R} packages for statistical learning include facilities for cross-validation. Notable examples are the \pkg{caret} package [@Kuhn:2008] and the \pkg{tidyfit} package [@Pfitzinger:2024]; the latter employs \pkg{rsample} [@FrickEtAl:2024] for CV, which we will discuss presently This contrasts with the approach taken in the \pkg{cv} package, which, as we have explained, directly supports classes of standard \proglang{R} statistical models.

Other packages, including \pkg{cvms} [@OlsenZachariae:2024], \pkg{mlexperiments} [@Kapsner:2024], \pkg{origami} [@CoyleEtAl:2022], and \pkg{rsample} (which also supports bootstrapping), modularize the CV process to a greater or lesser extent. Advantages of this approach include flexibility and generality, but a disadvantage is that use of these packages entails nontrivial programming effort and skill.

We illustrate with an example adapted from a vignette in the \pkg{rsample} package, which uses the \code{attrition} data set from the \pkg{modeldata} package. Here, with minimal explanation, is how one would perform LOO CV for this example using \pkg{rsample}:[^k-fold-used-in-rsample]
```{r rsample-example, cache=TRUE}
library(rsample)

data("attrition", package = "modeldata")
nrow(attrition)

# formula for model to be fit; Attrition is binary:
mod_form <- Attrition ~ JobSatisfaction + Gender + MonthlyIncome

rs_obj <- loo_cv(attrition) # original uses vfold_cv()

# user-supplied function:
holdout_results <- function(splits, ...) {
  mod <- glm(..., data=analysis(splits), family=binomial)
  holdout <- assessment(splits)
  res <- broom::augment(mod, newdata=holdout)
  lvls <- levels(holdout$Attrition)
  predictions <- factor(ifelse(res$.fitted > 0, lvls[2], lvls[1]),
                        levels=lvls)
  res$correct <- predictions == holdout$Attrition
  res
}

library(purrr, quietly=TRUE)
# apply CV:
rs_obj$results <- map(rs_obj$splits, holdout_results, mod_form)
# summarize results:
rs_obj$accuracy <- map_dbl(rs_obj$results,
                           function(x) mean(x$correct))
# accuracy is the complement of the Bayes-rule CV criterion:
1 - summary(rs_obj$accuracy) 
```
[^k-fold-used-in-rsample]: The example in the \pkg{rsample} vignette uses 10-fold rather than LOO CV, but otherwise the code from the vignette is only trivially modified here.

And here's how we can perform the same computation using the \code{cv()} function in the \pkg{cv} package:
```{r rsample-example-using-cv, cache=TRUE}
mod.attrition <- glm(mod_form, data=attrition, family=binomial)
(cv.attrition <- cv(mod.attrition, k="loo", criterion=BayesRule))
```

Let's compare the computational efficiency of the two implementations, also showing the use of \code{method="Woodbury"} and \code{method="hatvalues"} for \code{cv()}, which, in this example, produce numerically identical results to the default \code{method="exact"}:
```{r compare-rsample-cv, cache=TRUE}
all.equal(cv.attrition$`CV crit`,
          cv(mod.attrition, k="loo", criterion=BayesRule,
             method="Woodbury")$`CV crit`,
          )
all.equal(cv.attrition$`CV crit`,
          cv(mod.attrition, k="loo", criterion=BayesRule,
             method="hatvalues")$`CV crit`
          )

microbenchmark::microbenchmark(
  rsample={
    rs_obj$results <- map(rs_obj$splits, holdout_results, mod_form)
    rs_obj$accuracy <- map_dbl(rs_obj$results,
                               function(x) mean(x$correct))
  },
  cv.exact=cv(mod.attrition, k="loo", criterion=BayesRule),
  cv.wood=cv(mod.attrition, k="loo", criterion=BayesRule,
             method="Woodbury"),
  cv.hatvalues=cv(mod.attrition, k="loo", criterion=BayesRule,
                  method="hatvalues"),
  times=10
)
```
Computing time for \pkg{rsample} for this problem is similar to \code{cv()} with \code{method="exact"}, but more than an order of magnitude slower than \code{cv()} with \code{method="Woodbury"}, and more than three orders of magnitude slower then \code{cv()} with \code{method="hatvalues"}.

@@

Let $\mathrm{CV}(\mathbf{y}, \widehat{\mathbf{y}})$ represent a cross-validation cost criterion, such as mean-squared error, computed for all of the $n$ values of the response $\mathbf{y}$ based on fitted values $\widehat{\mathbf{y}}$ from the model fit to all of the data. We require that $\mathrm{CV}(\mathbf{y}, \widehat{\mathbf{y}})$ is the mean of casewise components, that is, $\mathrm{CV}(\mathbf{y}, \widehat{\mathbf{y}}) = \frac{1}{n}\sum_{i=1}^n\mathrm{cv}(y_i, \widehat{y}_i)$.[^contrast-function] For example, $\mathrm{MSE}(\mathbf{y}, \widehat{\mathbf{y}}) = \frac{1}{n}\sum_{i=1}^n (y_i - \widehat{y}_i)^2$.[^casewise-loss]

[^casewise-loss]: Some commonly employed CV criteria---such as the root-mean-squared error (RMSE), median absolute error, and, for binary-regression models, the complement of the area under the receiver operating characteristic (ROC) curve---are not means of casewise components. That the RMSE and median absolute error aren't means of casewise components is obvious; for the complement of the area under the ROC curve, see the vignette on extending the \pkg{cv} package.

### 2024-05-23 scratch/notes_john/JSS-article/JSS-article-3.tex
add section on other R CV packages to JSS paper

Residual standard error: 4.374 on 389 degrees of freedom
Multiple R-squared:  0.6876,    Adjusted R-squared:  0.686 
F-statistic:   428 on 2 and 389 DF,  p-value: < 2.2e-16
\end{CodeOutput}
\begin{CodeInput}
R> cv(m.auto, confint=TRUE)
\end{CodeInput}
\begin{CodeOutput}
R RNG seed set to 423102
\end{CodeOutput}
\begin{CodeOutput}
10-Fold Cross Validation
method: Woodbury
criterion: mse
cross-validation criterion = 19.27853
bias-adjusted cross-validation criterion = 19.26291
95% CI for bias-adjusted CV criterion = (15.80408, 22.72175)
full-sample criterion = 18.98477 
\end{CodeOutput}
\end{CodeChunk}

@@

\begin{CodeChunk}
\begin{CodeInput}
R> BayesRule
\end{CodeInput}
\begin{CodeOutput}
function (y, yhat) 
{
    if (!all(y %in% c(0, 1))) 
        stop("response values not all 0 or 1")
    if (any(yhat < 0) || any(yhat > 1)) 
        stop("fitted values outside of interval [0, 1]")
    yhat <- round(yhat)
    result <- mean(y != yhat)
    attr(result, "casewise loss") <- "y != round(yhat)"
    result
}
<bytecode: 0x12e288c00>
<environment: namespace:cv>
\end{CodeOutput}
\end{CodeChunk}

@@

@@

Let \(\mathrm{CV}(\mathbf{y}, \widehat{\mathbf{y}})\) represent a
cross-validation cost criterion, such as mean-squared error, computed
for all of the \(n\) values of the response \(\mathbf{y}\) based on
fitted values \(\widehat{\mathbf{y}}\) from the model fit to all of the
data. We require that \(\mathrm{CV}(\mathbf{y}, \widehat{\mathbf{y}})\)
is the mean of casewise components, that is,
\(\mathrm{CV}(\mathbf{y}, \widehat{\mathbf{y}}) = \frac{1}{n}\sum_{i=1}^n\mathrm{cv}(y_i, \widehat{y}_i)\).\footnote{\citet{ArlotCelisse:2010}
  term the casewise loss, \(\mathrm{cv}(y_i, \widehat{y}_i)\), the
  ``contrast function.''} For example,
\(\mathrm{MSE}(\mathbf{y}, \widehat{\mathbf{y}}) = \frac{1}{n}\sum_{i=1}^n (y_i - \widehat{y}_i)^2\).

>>>>

Residual standard error: 4.374 on 389 degrees of freedom
Multiple R-squared:  0.6876,    Adjusted R-squared:  0.686 
F-statistic:   428 on 2 and 389 DF,  p-value: < 2.2e-16
\end{CodeOutput}
\begin{CodeInput}
R> cv(m.auto, confint=TRUE)
\end{CodeInput}
\begin{CodeOutput}
R RNG seed set to 226376
\end{CodeOutput}
\begin{CodeOutput}
10-Fold Cross Validation
method: Woodbury
criterion: mse
cross-validation criterion = 19.33326
bias-adjusted cross-validation criterion = 19.31477
95% CI for bias-adjusted CV criterion = (15.83621, 22.79333)
full-sample criterion = 18.98477 
\end{CodeOutput}
\end{CodeChunk}

@@

\begin{CodeChunk}
\begin{CodeInput}
R> BayesRule
\end{CodeInput}
\begin{CodeOutput}
function(y, yhat) {
  if (!all(y %in% c(0, 1)))
    stop("response values not all 0 or 1")
  if (any(yhat < 0) ||
      any(yhat > 1))
    stop("fitted values outside of interval [0, 1]")
  yhat <- round(yhat)
  result <- mean(y != yhat) # proportion in error
  attr(result, "casewise loss") <- "y != round(yhat)"
  result
}
<bytecode: 0x14013a7f0>
<environment: namespace:cv>
\end{CodeOutput}
\end{CodeChunk}

@@

The \pkg{cv} package is far from unique in implementing cross-validation
of regression models in \proglang{R}. We've already mentioned the
\code{cv.glm()} function in the \pkg{boot} package. A general review of
CV facilities in \proglang{R} is beyond the scope of the current paper,
but in this section we remark selectively on other \proglang{R} packages
that support CV.

Our goal in implementing CV is to make it conveniently available to
\proglang{R} users with limited programming skills,\footnote{\proglang{R}
  users with sophisticated programming skills would generally find it
  unchallenging to implement CV directly for specific applications. Even
  in this case, however, there's an argument for using a simple
  pre-programmed interface to CV, to minimize programming effort and to
  avoid mistakes.} and hence to encourage its use. Towards this end, the
\pkg{cv} package provides a simple, consistent interface through the
\code{cv()} generic function, with specific methods for different
classes of standard \proglang{R} statistical models. Moreover, we have
tried to make writing methods for additional classes of statistical
models as simple and straightforward as possible. Because CV is
computationally intensive, we also aim for efficiency, by enabling
parallel computations generally, and by exploiting properties of
specific classes of statistical models (in particular, computations
based on hatvalues and the Woodbury matrix identity for linear and
generalized linear model). Finally, the package includes some unusual
features, such as general support for mixed-effects models and for
cross-validating complex model-specification procedures.

Some \proglang{R} packages for statistical learning include facilities
for cross-validation. Notable examples are the \pkg{caret} package
\citep{Kuhn:2008} and the \pkg{tidyfit} package \citep{Pfitzinger:2024};
the latter employs \pkg{rsample} \citep{FrickEtAl:2024} for CV, which we
will discuss presently This contrasts with the approach taken in the
\pkg{cv} package, which, as we have explained, directly supports classes
of standard \proglang{R} statistical models.

Other packages, including \pkg{cvms} \citep{OlsenZachariae:2024},
\pkg{mlexperiments} \citep{Kapsner:2024}, \pkg{origami}
\citep{CoyleEtAl:2022}, and \pkg{rsample} (which also supports
bootstrapping), modularize the CV process to a greater or lesser extent.
Advantages of this approach include flexibility and generality, but a
disadvantage is that use of these packages entails nontrivial
programming effort and skill.

We illustrate with an example adapted from a vignette in the
\pkg{rsample} package, which uses the \code{attrition} data set from the
\pkg{modeldata} package. Here, with minimal explanation, is how one
would perform LOO CV for the example using \pkg{rsample}:\footnote{The
  example in the \pkg{rsample} vignette uses 10-fold rather than LOO CV,
  but otherwise the code from the vignette is only trivially modified
  here.}

\begin{CodeChunk}
\begin{CodeInput}
R> library(rsample)
R> 
R> data("attrition", package = "modeldata")
R> nrow(attrition)
\end{CodeInput}
\begin{CodeOutput}
[1] 1470
\end{CodeOutput}
\begin{CodeInput}
R> # formula for model to be fit; Attrition is binary:
R> mod_form <- Attrition ~ JobSatisfaction + Gender + MonthlyIncome
R> 
R> rs_obj <- loo_cv(attrition) # original uses vfold_cv()
R> 
R> # user-supplied function:
R> holdout_results <- function(splits, ...) {
+   mod <- glm(..., data=analysis(splits), family=binomial)
+   holdout <- assessment(splits)
+   res <- broom::augment(mod, newdata=holdout)
+   lvls <- levels(holdout$Attrition)
+   predictions <- factor(ifelse(res$.fitted > 0, lvls[2], lvls[1]),
+                         levels=lvls)
+   res$correct <- predictions == holdout$Attrition
+   res
+ }
R> 
R> library(purrr, quietly=TRUE)
R> # apply CV:
R> rs_obj$results <- map(rs_obj$splits, holdout_results, mod_form)
R> # summarize results:
R> rs_obj$accuracy <- map_dbl(rs_obj$results,
+                            function(x) mean(x$correct))
R> # accuracy is the complement of the Bayes-rule CV criterion:
R> 1 - summary(rs_obj$accuracy) 
\end{CodeInput}
\begin{CodeOutput}
   Min. 1st Qu.  Median    Mean 3rd Qu.    Max. 
 1.0000  0.0000  0.0000  0.1612  0.0000  0.0000 
\end{CodeOutput}
\end{CodeChunk}

And here's how we can perform the same computation using the \code{cv()}
function in the \pkg{cv} package:

\begin{CodeChunk}
\begin{CodeInput}
R> mod.attrition <- glm(mod_form, data=attrition, family=binomial)
R> (cv.attrition <- cv(mod.attrition, k="loo", criterion=BayesRule))
\end{CodeInput}
\begin{CodeOutput}
n-Fold Cross Validation
method: exact
criterion: BayesRule
cross-validation criterion = 0.1612245
bias-adjusted cross-validation criterion = 0.1612245
95% CI for bias-adjusted CV criterion = (0.1424194, 0.1800296)
full-sample criterion = 0.1612245 
\end{CodeOutput}
\end{CodeChunk}

Let's compare the computational efficiency of the two implementations,
also showing the use of \code{method="Woodbury"} and
\code{method="hatvalues"} for \code{cv()}, which, in this example,
produce numerically identical results to the default
\code{method="exact"}:

\begin{CodeChunk}
\begin{CodeInput}
R> all.equal(cv.attrition$`CV crit`,
+           cv(mod.attrition, k="loo", criterion=BayesRule,
+              method="Woodbury")$`CV crit`,
+           )
\end{CodeInput}
\begin{CodeOutput}
[1] TRUE
\end{CodeOutput}
\begin{CodeInput}
R> all.equal(cv.attrition$`CV crit`,
+           cv(mod.attrition, k="loo", criterion=BayesRule,
+              method="hatvalues")$`CV crit`
+           )
\end{CodeInput}
\begin{CodeOutput}
[1] TRUE
\end{CodeOutput}
\begin{CodeInput}
R> microbenchmark::microbenchmark(
+   rsample={
+     rs_obj$results <- map(rs_obj$splits, holdout_results, mod_form)
+     rs_obj$accuracy <- map_dbl(rs_obj$results,
+                                function(x) mean(x$correct))
+   },
+   cv.exact=cv(mod.attrition, k="loo", criterion=BayesRule),
+   cv.wood=cv(mod.attrition, k="loo", criterion=BayesRule,
+              method="Woodbury"),
+   cv.hatvalues=cv(mod.attrition, k="loo", criterion=BayesRule,
+                   method="hatvalues"),
+   times=10
+ )
\end{CodeInput}
\begin{CodeOutput}
Warning in microbenchmark::microbenchmark(rsample = {: less accurate nanosecond
times to avoid potential integer overflows
\end{CodeOutput}
\begin{CodeOutput}
Unit: milliseconds
         expr         min          lq        mean      median          uq
      rsample 5814.989369 5864.285227 5952.773502 5897.248509 6023.438043
     cv.exact 4739.827837 4827.064193 4902.546681 4910.249278 4952.683150
      cv.wood  103.663047  104.760125  110.628578  111.201532  116.027007
 cv.hatvalues    1.217905    1.222251    1.409719    1.479649    1.543937
         max neval
 6207.205988    10
 5090.904318    10
  118.610540    10
    1.591538    10
\end{CodeOutput}
\end{CodeChunk}

Computing time for \pkg{rsample} for this problem is similar to
\code{cv()} with \code{method="exact"}, but more than an order of
magitude slower than \code{cv()} with \code{method="Woodbury"}, and more
than three orders of magnitude slower then \code{cv()} with
\code{method="hatvalues"}.

@@

Let \(\mathrm{CV}(\mathbf{y}, \widehat{\mathbf{y}})\) represent a
cross-validation cost criterion, such as mean-squared error, computed
for all of the \(n\) values of the response \(\mathbf{y}\) based on
fitted values \(\widehat{\mathbf{y}}\) from the model fit to all of the
data. We require that \(\mathrm{CV}(\mathbf{y}, \widehat{\mathbf{y}})\)
is the mean of casewise components, that is,
\(\mathrm{CV}(\mathbf{y}, \widehat{\mathbf{y}}) = \frac{1}{n}\sum_{i=1}^n\mathrm{cv}(y_i, \widehat{y}_i)\).\footnote{\citet{ArlotCelisse:2010}
  term the casewise loss, \(\mathrm{cv}(y_i, \widehat{y}_i)\), the
  ``contrast function.''} For example,
\(\mathrm{MSE}(\mathbf{y}, \widehat{\mathbf{y}}) = \frac{1}{n}\sum_{i=1}^n (y_i - \widehat{y}_i)^2\).\footnote{Some
  commonly employed CV criteria---such as the root-mean-squared error
  (RMSE), median absolute error, and, for binary-regression models, the
  complement of the area under the receiver operating characteristic
  (ROC) curve---are not means of casewise components. That the RMSE and
  median absolute error aren't means of casewise components is obvious;
  for the complement of the area under the ROC curve, see the vignette
  on extending the \pkg{cv} package.}
